Predicting instabilities of a tuneable ring laser with an iterative map model

### 2021-02-21 Papers/Journal/Tuneable_Lasers.tex
Final read through

\begin{abstract}
	Simple mathematical models have been unable to predict the conditions leading to instabilities in a tuneable ring laser. Here, we propose a nonlinear iterative map model for tuneable ring lasers. Solving a reduced nonlinear Schr\"odinger equation for each component in the laser cavity, we obtain an algebraic map for each component. Iterating through the maps gives the total effect of one round trip. By neglecting the nonlinearity, we find a linearly chirped Gaussian to be the analytic fixed point solution, which we analyze asymptotically. We then numerically solve the full nonlinear model, allowing us to probe the underlying interplay of dispersion, modulation, and nonlinearity, as the pulse evolves over hundreds of round trips of the cavity. In the nonlinear case, we find the chirp saturates, and the Fourier transform of the pulse becomes more rectangular in shape. Finally, for a nominal plane in the parameter space, we uncover a rich, sharp boundary separating the stable region and the unstable region where modulation instability degrades the pulse into an unsustainable state.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Dispersion-tuned actively mode-locked lasers are an active area of optics research since they have a wide range of applications. These lasers can have ultrashort pulses, as short as 50 femtoseconds \cite{chung2017}, and the ability to tune their output frequency rapidly over a range of over 100 nm~\cite{bohun2015, burgoyne2010, chung2017, yamashita2009}. The capability to alter the operating frequency makes tuneable lasers particularly useful for imaging applications, such as optical coherence tomography~\cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy~\cite{burgoyne2014}, and deep tissue multi-photon microscopy~\cite{chung2017}, as well as sensing and measuring of other ultrafast processes~\cite{burgoyne2014, silfvast2004, oktem2010}. Tuneable lasers are usually constructed in a ring with five main components: an optical coupler, chirped fibre Bragg grating (CFBG), modulator (or saturable absorber), gain fibre (usually doped with Er), and pump laser. A typical ring laser cavity is depicted in Fig.~\ref{fig:cavity}. The Kerr nonlinearity plays a crucial role in the dynamics within the cavity due to the high power and ultrashort duration of the pulses.

Several effects arise within the cavity due to the interplay of dispersion, modulation, and the nonlinearity~\cite{bohun2015, coen1997, lapre2019, meng2020, oktem2010, shao2019, woodward2018}. The two effects of most interest for us are wave breaking, and modulation instability. Wave breaking causes the leading edge of a pulse to redshift, and the trailing edge to blueshift (the effect reverses in the anomalous dispersion regime) to the point a shock develops~\cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The frequency shift causes the pulse to become more rectangular in the frequency domain, with a linear chirp over most of the pulse, saturating at the leading and trailing edge. In optics, wave breaking arises from self-phase modulation (SPM), which is a direct effect of the Kerr nonlinearity, causing the pulse to interfere with itself. New frequencies are then generated such that a rectangular profile is induced~\cite{agrawal2013, woodward2018}. Modulation instability---which typically arises in the anomalous dispersion regime---is the other effect that will be of interest. However, in the presence of multiple frequencies, modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation and four-wave mixing~\cite{agrawal1987, agrawal2013, haelterman1992}. Modulation instability causes a pulse to degrade into multiple much shorter, less intense, pulses and can cause a laser to lose coherence~\cite{agrawal1987, coen1997, haelterman1992}. Outside of optics, wave breaking and modulation instability occur in other areas with nonlinear waves and dispersive media such as plasma physics, heat propagation, transmission lines, and fluid dynamics~\cite{coen1997, rothenberg1989b}. Within a ring laser cavity, wave breaking and modulation instability can become parasitic leading to an unstable and unsustainable pulse. Understanding the rich landscape of the parameter space is important for determining design principles to guarantee the ring laser is stable and sustainable~\cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

\begin{figure}[tbp]
	\centering
	\input{Cavity}
	\caption{Common cavity of a fibre ring laser~\cite{burgoyne2014, chung2017, lapre2019, shao2019}. The pump laser travels clockwise around the left-hand loop (red arrows) energizing the Er-doped gain fibre. The main laser travels clockwise around the right-hand loop (blue arrows), with most of the pulse exiting the cavity through the optical coupler.}
	\label{fig:cavity}
\end{figure}

@@

The GNLSE, \eqref{eq:nlse}, has been applied in nonlinear optics~\cite{agrawal2013} and fibre optic communications~\cite{agrawal2002}. However, \eqref{eq:nlse} is regularly generalized even further to include a modulation term. The resulting equation is referred to as the master equation of mode-locking (see \cite{haus1975, haus1984, haus2000, tamura1996, usechak2005}), or sometimes as the Ginzburg--Landau equation. Owing to the nonlinear term in \eqref{eq:nlse} and the often non-trivial form of $g(A)$, as well as the possible addition of a modulation term, no analytic solution is known to the master equation or to \eqref{eq:nlse}. A comprehensive description of the theory and history of mode-locking lasers is presented in Haus~\cite{haus2000}. While the GNLSE well represents the dynamics in, say, a single optical fibre, \eqref{eq:nlse} is not entirely representative of the underlying physics within a laser cavity such as Fig. \ref{fig:cavity}. The derivation of \eqref{eq:nlse} assumes that each process affects the pulse continuously within the cavity. Therefore, the GNLSE, and generalizations, can be thought of as an `average' model for lasers---the assumption of continuously effecting the pulse breaks down if the pulse undergoes large changes during any one round trip. Often, this assumption does not hold because of the specialized, discrete components.

To resolve some of the issues with an `average' model, each component and region in the laser is modelled with a separate GNLSE. The solution of one GNLSE is used as the initial condition for the GNLSE for the next component~\cite{lapre2019, meng2020, oktem2010, woodward2018}. An iterative map model allows for better encapsulation of specific dynamics introduced by each component. The main drawback of this method is the extreme computational expense since a nonlinear equation must be numerically solved for every segment for each of the hundreds of round trips in a given simulation. We can alleviate this computational expense by noticing each effect is localized to a corresponding component in the cavity: a majority of the loss is due to the optical coupler, virtually all of the dispersion happens within the CFBG~\cite{agrawal2002}, the pulse is only modulated by the modulator, and, the pulse is amplified only within the Er-doped gain fibre. Therefore, we can drastically simplify each of the GNLSEs into a form we can solve analytically. Using analytic maps instead of numerically solving a GNLSE reduces the computational cost, while maintaining the benefit of reflecting the specific geometry of a component-wise simulation.

@@

\section{Iterative Map Model}
\label{sec:model}
We derive our own iterative map model for the evolution of a pulse through a laser cavity by solving a reduced GNLSE for each component in the cavity---except modulation, which we consider to be applied externally by an electro-optic modulator. Starting with the gain component, we reduce \eqref{eq:nlse} to 
\begin{align}
	\pdiff{A}{z} &= \frac{1}{2}g(A)A,
	\label{eq:gainexample}
\end{align}
by assuming all other effects are negligible. Once again, our assumption is justified by the localization of each effect to its corresponding component, allowing us to simplify \eqref{eq:nlse} to \eqref{eq:gainexample} in the gain fibre. To proceed, we assume the gain has the form
\begin{align}
	g(A) &= \frac{g_0}{1 + E / \Es},
	\label{eq:gainform}
\end{align}
where $g_0$ is a small signal gain, $E$ is the energy of the pulse, defined as
\begin{align}
	E &= \int_{-\infty}^\infty |A|^2 \, \df T,
	\label{eq:energy}
\end{align}
and $\Es$ is the energy that the gain begins to saturate~\cite{haus1984, shtyrina2017, silfvast2004}, which depends on the power of the pump laser. With the choice of $g(A)$ given in \eqref{eq:gainform}, \eqref{eq:gainexample} becomes
\begin{align}
	\pdiff{A}{z} &= \frac{g_0 A}{2 \left( 1 + E / \Es \right)}.
	\label{eq:pdiffgain}
\end{align}
Multiplying \eqref{eq:pdiffgain} by the complex conjugate of $A$, we have
\begin{align}
	\overline{A} \pdiff{A}{z} &= \frac{g_0}{2 \left( 1 + E / \Es \right)} |A|^2.
	\label{eq:pdiffgain2}
\end{align}
By adding \eqref{eq:pdiffgain2} and the complex conjugate of \eqref{eq:pdiffgain2}, and then integrating over the pulse, we have
\begin{align}
	\diff{E}{z} &= \frac{g_0 E}{1 + E / \Es}.
	\label{eq:energyde}
\end{align}
By assuming the energy entering the gain fibre is $E_g$ and the energy after travelling through the gain fibre of length $L_g$ is $E_{\text{out}}$, we integrate \eqref{eq:energyde} to find the energy after amplification is
\begin{align}
	\label{eq:engain}
	E_{\text{out}} &= \Es W \left( \frac{E_g}{\Es} \exp \left( \frac{E_g}{\Es} \right) \exp( g_0 L_g ) \right),
\end{align}
where $W$ is the Lambert $W$ function. Moreover, as the gain will amplify the pulse without changing the shape of the envelope, the pulse after travelling through the gain fibre can be expressed as
\begin{align}
	G(A) &= \mathcal{G} A.
	\label{eq:ampfact}
\end{align}
Then,
\begin{align}
	\mathcal{G}^2 E_g &= E_\text{out}
\end{align}
by \eqref{eq:energy}, \eqref{eq:engain} and \eqref{eq:ampfact}, therefore,
\begin{align}
	G(A) &= \left( \frac{E_\text{out}}{E_g} \right)^{1/2} A, \\
	&=	\left( \frac{\Es}{E_g} W \left( \frac{E_g}{\Es} \exp \left( \frac{E_g}{\Es} \right) \exp( g_0 L_g ) \right) \right)^{1/2} A \label{eq:dimgain}
\end{align}
for the amplification of the pulse after the Er-doped gain fibre.

@@

Finally, we consider the modulation. Here, we assume the pulse is modulated by an electro-optic modulator. For simplicity, the form is taken as the Gaussian
\begin{align}
	M(A) &= \exp( -T^2 / 2 T_M^2 ) A,
	\label{eq:modform}
\end{align}
where $T_M$ is the characteristic width of the modulation. This assumption is not restrictive. Calcaterra and Boldt~\cite{calcaterra2008a} have shown a sum of Gaussian functions can approximate any square integrable function. Therefore, our solution extends to any continuous, bounded modulation function. Additionally, using a saturable absorber in place of a modulator is common in practice. In this case, we can exchange \eqref{eq:modform} for~\cite{lapre2019, meng2020, oktem2010, woodward2018}
\begin{align}
	S(A) &= \left( 1 - \frac{q_0}{1 + |A|^2 / P_\text{sat}} \right) A,
\end{align}
where $q_0$ is the modulation depth, $|A|^2$ is the instantaneous power of the pulse, and $P_\text{sat}$ is the saturation power.

@@

\begin{figure}[tbp]
	\centering
	\includegraphics{Figures/Evo}
	\caption{Evolution of the envelope during one round trip of the cavity at equilibrium. The pulse decays because of the optical coupler, disperses because of the CFBG, is modulated by the modulator, and finally, is amplified by the gain fibre (the envelope is unaltered by the nonlinearity). Note, the $z$ axis is not to scale.}
	\label{fig:cavityevo}
\end{figure}

@@

\subsection{Non-Dimensionalization}
Non-dimensionalization is a technique used in mathematical modelling to better understand the relative importance of different processes within a system. In essence, we scale all variables by typical values, based on the geometry and boundary conditions (see, for example,~\cite{howison2005}). Doing so `factors' the dimensional units out of the problem---leaving only dimensionless equations with dimensionless parameters. The relative magnitudes of the dimensionless parameters characterize the dominant dynamics---like how the Reynolds number describes fluid flow. Each process in the laser can be better understood by re-scaling the time, energies, and amplitude by the convenient scalings:
\begin{align}
	T &= T_M \widetilde{T},& E &= \Es \widetilde{E},& A &= \left( \frac{\Es}{T_M} \right)^{1/2} \widetilde{A}.
\end{align}

@@

We consider the pulse given by
\begin{align}
	A = \sqrt{P} \exp \left( -(1 + iC) \frac{T^2}{2 \sigma^2} \right) \exp(i \phi_0),
	\label{eq:A0}
\end{align}
where $P$ is the peak power, $C$ is the chirp, $\sigma^2$ is the variance, and $\phi_0$ is the phase; the pulse is shown in Fig.~\ref{fig:samplegauss}. By computing \eqref{eq:A0} after one round trip, and solving $\mathcal{L}(A) = A \textrm{e}^{i \phi}$, tedious manipulations yield the conditions
\begin{align}
	\frac{\sigma^4}{1 - \sigma^2} &= (\sigma^2 + 2 C s^2)^2 + 4 s^4, \label{eq:sigrelation} \\
	\frac{C}{1 - \sigma^2} &= C + 2 \frac{s^2}{\sigma^2} (1 + C^2), \label{eq:chirprelation} \\
	1 &= \frac{W(a E_g \textrm{e}^{E_g})}{E_g} h^2 (1 - \sigma^2)^{1/2} \label{eq:energyrelation}
\end{align}
for \eqref{eq:A0} to be in equilibrium. Using \eqref{eq:sigrelation} and \eqref{eq:chirprelation} we eliminate the chirp, $C$, and find an expression relating $\sigma^2$ and $s$, namely
\begin{align}
	\sigma^8 + 4 s^4 \sigma^6 - 20 s^4 \sigma^4 + 32 s^4 \sigma^2 - 16 s^4 = 0.
	\label{eq:var}
\end{align}
\eqref{eq:var} has the positive, real, analytic solution
\begin{equation}
	\begin{split}
		\sigma^2 = \sqrt{2} s \left( s^6 + 3s^2 + \sqrt{4 + s^4} \left( 1 + s^4 \right) \right)^{1/2} & \\
		- s^4 - s^2 & \sqrt{4 + s^4}.
		\label{eq:equilvar}
	\end{split}
\end{equation}
Substituting \eqref{eq:equilvar} into \eqref{eq:chirprelation}, we find the chirp is
\begin{align}
	C = \frac{\sigma^4 - \left( \sigma^8 - 16 s^4 \left( 1 - \sigma^2 \right)^2 \right)^{1/2}}{4 s^2 \left( 1 - \sigma^2 \right)}.
	\label{eq:chirp}
\end{align}
We obtain the energy of the pulse entering the gain fibre, $E_g$, from \eqref{eq:energyrelation}:
\begin{align}
	E_g &= \frac{h^2 \left( 1 - \sigma^2 \right)^{1/2} \log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{h^2 \left( 1 - \sigma^2 \right)^{1/2} - 1}
\end{align}
and, thus, the equilibrium energy at the optical coupler, is
\begin{align}
	E &= W \left( a E_g \textrm{e}^{E_g} \right) \\
	&= \frac{\log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{1 - h^2 \left( 1 - \sigma^2 \right)^{1/2}}.
	\label{eq:analenergy}
\end{align}
The argument of the logarithm in \eqref{eq:analenergy} gives us the condition
\begin{align}
	a h^2 (1 - \sigma^2)^{1/2} > 1
	\label{eq:energycond}
\end{align}
for the pulse to be sustainable.

Equating the energy of the pulse in \eqref{eq:A0} and \eqref{eq:analenergy}, we find the peak power to be
\begin{align}
	P &= \frac{\log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{\sqrt{\pi} \sigma \left( 1 - h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}.
	\label{eq:analpower}
\end{align}

The form of \eqref{eq:equilvar}--\eqref{eq:analpower} make the expressions difficult to work with. However, recall that $s$ is small, thus, we asymptotically expand \eqref{eq:equilvar}--\eqref{eq:analpower} assuming $s \ll 1$, and
\begin{align}
	\sigma^2 &\sim 2s(1 - s) + \bigO{s^3}, \\
	C &\sim 1 - s + \frac{1}{2}s^2 + \bigO{s^3}, \\
	E &\sim \frac{\log (a h^2)}{1 - h^2} - \frac{1}{1 - h^2} \left( 1 + \frac{h^2 \log(a h^2)}{1 - h^2}  \right) s + \bigO{s^2}, \label{eq:maxenergy} \\
	P &\sim \frac{\log(ah^2)}{\sqrt{2 \pi}(1 - h^2)} s^{-1/2} + \bigO{s^{1/2}},
\end{align}
in the limit $s \rightarrow 0$. The expansions in the limit $s \rightarrow \infty$ are given in Appendix \ref{sec:largedisp}.

@@

Fig.~\ref{fig:breakevo} highlights the phenomenon of modulation instability. By decreasing the dispersion from Fig.~\ref{fig:chirpft}, we find the pulse begins to `breathe'. During the first two dozen round trips of the cavity the SPM compounds and becomes detrimental to the pulse. In turn, modulation instability is induced, and degrades the pulse, until the pulse is no longer stable or sustainable. On the other hand, having a more moderate value of $b$, the pulse is able to equilibrate even with a less favourable, random, initial pulse (Fig.~\ref{fig:convevo}), as in~\cite{meng2020}. The pulse is able to shed the initial left lobe ($T \approx -1 / 2$) because of dispersion and modulation, and a central lobe ($T \approx 0$) forms, which is able to grow because of the gain medium. The central lobe then allows the pulse to come to equilibrium quickly. Furthermore, since the intensity of the initial pulse is small relative to the saturation energy, the effect of SPM is negligible during the first few round trips, allowing the laser to select preferable modes. The preferable modes then get amplified, stabilizing the pulse. Evidently, the initial shape of the pulse is much less important than the initial energy of the pulse, or the strength of the nonlinearity.

>>>>

\begin{abstract}
	Simple mathematical models have been unable to predict the conditions leading to instabilities in a tuneable ring laser. Here, we propose a nonlinear iterative map model for tuneable ring lasers. Solving a reduced nonlinear Schr\"odinger equation for each component in the laser cavity, we obtain an algebraic map for each component. Iterating through the maps gives the total effect of one round trip. By neglecting the nonlinearity, we find a linearly chirped Gaussian to be the analytic fixed point solution, which we analyze asymptotically. We then numerically solve the full nonlinear model, allowing us to probe the underlying interplay of dispersion, modulation, and nonlinearity as the pulse evolves over hundreds of round trips of the cavity. In the nonlinear case, we find the chirp saturates, and the Fourier transform of the pulse becomes more rectangular in shape. Finally, for a nominal plane in the parameter space, we uncover a rich, sharp boundary separating the stable region and the unstable region where modulation instability degrades the pulse into an unsustainable state.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Dispersion-tuned actively mode-locked lasers are an active area of optics research since they have a wide range of applications. These lasers can have ultrashort pulses, as short as 50 femtoseconds \cite{chung2017}, and the ability to tune their output frequency rapidly over a range of over 100 nm~\cite{bohun2015, burgoyne2010, chung2017, yamashita2009}. The capability to alter the operating frequency makes tuneable lasers particularly useful for imaging applications, such as optical coherence tomography~\cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy~\cite{burgoyne2014}, and deep tissue multi-photon microscopy~\cite{chung2017}, as well as sensing and measuring of other ultrafast processes~\cite{burgoyne2014, silfvast2004, oktem2010}. Tuneable lasers are usually constructed in a ring with five main components: an optical coupler, a chirped fibre Bragg grating (CFBG), a modulator (or saturable absorber), a gain fibre (usually doped with Er), and a pump laser. A typical ring laser cavity is depicted in Fig.~\ref{fig:cavity}. The Kerr nonlinearity plays a crucial role in the dynamics within the cavity due to the high power and ultrashort duration of the pulses.

Several effects arise within the cavity due to the interplay of dispersion, modulation, and the nonlinearity~\cite{bohun2015, coen1997, lapre2019, meng2020, oktem2010, shao2019, woodward2018}. The two effects of most interest for us are wave breaking, and modulation instability. Wave breaking causes the leading edge of a pulse to redshift, and the trailing edge to blueshift (the effect reverses in the anomalous dispersion regime) to the point a shock develops~\cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The frequency shift causes the pulse to become more rectangular in the frequency domain, with a linear chirp over most of the pulse, saturating at the leading and trailing edge. In optics, wave breaking arises from self-phase modulation (SPM), which is a direct effect of the Kerr nonlinearity, causing the pulse to interfere with itself. New frequencies are then generated such that a rectangular profile is induced~\cite{agrawal2013, woodward2018}. Modulation instability---which typically arises in the anomalous dispersion regime---is the other effect that will be of interest. However, in the presence of multiple frequencies, modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation and four-wave mixing~\cite{agrawal1987, agrawal2013, haelterman1992}. Modulation instability causes a pulse to degrade into multiple, much shorter, less intense pulses and can cause a laser to lose coherence~\cite{agrawal1987, coen1997, haelterman1992}. Outside of optics, wave breaking and modulation instability occur in other areas with nonlinear waves and dispersive media such as plasma physics, heat propagation, transmission lines, and fluid dynamics~\cite{coen1997, rothenberg1989b}. Within a ring laser cavity, wave breaking and modulation instability can become parasitic, leading to an unstable and unsustainable pulse. Understanding the rich landscape of the parameter space is important for determining design principles to guarantee the ring laser is stable and sustainable~\cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

\begin{figure}[tbp]
	\centering
	\input{Cavity}
	\caption{Common cavity construction of a fibre ring laser~\cite{burgoyne2014, chung2017, lapre2019, shao2019}. The pump laser travels clockwise around the left-hand loop (red arrows) energizing the Er-doped gain fibre. The main laser travels clockwise around the right-hand loop (blue arrows), with most of the pulse exiting the cavity through the optical coupler.}
	\label{fig:cavity}
\end{figure}

@@

The GNLSE, \eqref{eq:nlse}, has been applied in nonlinear optics~\cite{agrawal2013} and fibre optic communications~\cite{agrawal2002}. However, \eqref{eq:nlse} is regularly generalized even further to include a modulation term. The resulting equation is referred to as the master equation of mode-locking (see \cite{haus1975, haus1984, haus2000, tamura1996, usechak2005}), or sometimes as the Ginzburg--Landau equation. Owing to the nonlinear term in \eqref{eq:nlse} and the often non-trivial form of $g(A)$, as well as the possible addition of a modulation term, no analytic solution is known to the master equation or to \eqref{eq:nlse}. A comprehensive description of the theory and history of mode-locking lasers is presented in Haus~\cite{haus2000}. While the GNLSE well represents the dynamics in, say, a single optical fibre, \eqref{eq:nlse} is not entirely representative of the underlying physics within a laser cavity such as Fig. \ref{fig:cavity}. The derivation of \eqref{eq:nlse} assumes that each process affects the pulse continuously within the cavity. Therefore, the GNLSE, and generalizations, can be thought of as an `average' model for lasers---the assumption of continuously effecting the pulse breaks down if the pulse undergoes large changes during any one round trip of the cavity. Often, this assumption does not hold because of the specialized, discrete components.

To resolve some of the issues with an `average' model, each component and region in the laser is modelled with a separate GNLSE. The solution of one GNLSE is used as the initial condition for the GNLSE for the next component~\cite{lapre2019, meng2020, oktem2010, woodward2018}. An iterative map model allows for better encapsulation of specific dynamics introduced by each component. The main drawback of this method is the extreme computational expense since a nonlinear partial differential equation must be numerically solved for every segment for each of the hundreds of round trips in a given simulation. We can alleviate this computational expense by noticing each effect is localized to a corresponding component in the cavity: a majority of the loss is due to the optical coupler, virtually all of the dispersion happens within the CFBG~\cite{agrawal2002}, the pulse is only modulated by the modulator, and the pulse is amplified only within the Er-doped gain fibre. Therefore, we can drastically simplify each of the GNLSEs into a form we can solve analytically. Using analytic maps instead of numerically solving a GNLSE reduces the computational cost, while maintaining the benefit of reflecting the specific geometry of a component-wise simulation.

@@

\section{Iterative Map Model}
\label{sec:model}
We derive our own iterative map model for the evolution of a pulse through a laser cavity by solving a reduced GNLSE for each component in the cavity---except modulation, which we consider to be applied externally by an electro-optic modulator. Starting with the gain component, we reduce \eqref{eq:nlse} to 
\begin{align}
	\pdiff{A}{z} &= \frac{1}{2}g(A)A,
	\label{eq:gainexample}
\end{align}
by assuming all other effects are negligible. Once again, our assumption is justified by the localization of each effect to its corresponding component, allowing us to simplify \eqref{eq:nlse} to \eqref{eq:gainexample} in the gain fibre. To proceed, we assume the gain has the form
\begin{align}
	g(A) &= \frac{g_0}{1 + E / \Es},
	\label{eq:gainform}
\end{align}
where $g_0$ is a small signal gain, $E$ is the energy of the pulse, defined as
\begin{align}
	E &= \int_{-\infty}^\infty |A|^2 \, \df T,
	\label{eq:energy}
\end{align}
and $\Es$ is the energy that the gain begins to saturate~\cite{haus1984, shtyrina2017, silfvast2004}, which depends on the power of the pump laser. With the choice of $g(A)$ given in \eqref{eq:gainform}, \eqref{eq:gainexample} becomes
\begin{align}
	\pdiff{A}{z} &= \frac{g_0 A}{2 \left( 1 + E / \Es \right)}.
	\label{eq:pdiffgain}
\end{align}
Multiplying \eqref{eq:pdiffgain} by the complex conjugate of $A$, we have
\begin{align}
	\overline{A} \pdiff{A}{z} &= \frac{g_0}{2 \left( 1 + E / \Es \right)} |A|^2.
	\label{eq:pdiffgain2}
\end{align}
By adding \eqref{eq:pdiffgain2} and the complex conjugate of \eqref{eq:pdiffgain2}, and then integrating over $T$, we have
\begin{align}
	\diff{E}{z} &= \frac{g_0 E}{1 + E / \Es}.
	\label{eq:energyde}
\end{align}
By assuming the energy entering the gain fibre is $E_g$ and the energy after travelling through the gain fibre of length $L_g$ is $E_{\text{out}}$, we integrate \eqref{eq:energyde} to find the energy after amplification is
\begin{align}
	\label{eq:engain}
	E_{\text{out}} &= \Es W \left( \frac{E_g}{\Es} \exp \left( \frac{E_g}{\Es} \right) \exp( g_0 L_g ) \right),
\end{align}
where $W$ is the Lambert $W$ function. Moreover, as the gain will amplify the pulse without changing the shape of the envelope, the pulse after travelling through the gain fibre can be expressed as
\begin{align}
	G(A) &= \mathcal{G} A.
	\label{eq:ampfact}
\end{align}
Then,
\begin{align}
	\mathcal{G}^2 E_g &= E_\text{out}
\end{align}
by \eqref{eq:energy}, \eqref{eq:engain} and \eqref{eq:ampfact}, therefore,
\begin{align}
	G(A) &= \left( \frac{E_\text{out}}{E_g} \right)^{1/2} A, \\
	&=	\left( \frac{\Es}{E_g} W \left( \frac{E_g}{\Es} \exp \left( \frac{E_g}{\Es} \right) \exp( g_0 L_g ) \right) \right)^{1/2} A \label{eq:dimgain}
\end{align}
for the amplification of the pulse after the Er-doped gain fibre.

@@

Finally, we consider the modulation. Here, we assume the pulse is modulated by an electro-optic modulator. For simplicity, the form is taken as the Gaussian
\begin{align}
	M(A) &= \exp( -T^2 / 2 T_M^2 ) A,
	\label{eq:modform}
\end{align}
where $T_M$ is the characteristic width of modulation. This assumption is not restrictive. Calcaterra and Boldt~\cite{calcaterra2008a} have shown a sum of Gaussian functions can approximate any square integrable function. Therefore, our solution extends to any continuous, bounded modulation function. Additionally, using a saturable absorber in place of an electro-optic modulator is common in practice. In this case, we can exchange \eqref{eq:modform} for~\cite{lapre2019, meng2020, oktem2010, woodward2018}
\begin{align}
	S(A) &= \left( 1 - \frac{q_0}{1 + |A|^2 / P_\text{sat}} \right) A,
\end{align}
where $q_0$ is the modulation depth, and $P_\text{sat}$ is the saturation power.

@@

\begin{figure}[tbp]
	\centering
	\includegraphics{Figures/Evo}
	\caption{Evolution of the envelope during one round trip of the cavity. The pulse decays because of the optical coupler, disperses because of the CFBG, is modulated by the modulator, and finally, is amplified by the gain fibre (the envelope is unaltered by the nonlinearity). Note, the $z$ axis is not to scale.}
	\label{fig:cavityevo}
\end{figure}

@@

\subsection{Non-Dimensionalization}
Non-dimensionalization is a technique used in mathematical modelling to better understand the relative importance of different processes within a system. In essence, we scale all variables by typical values, based on the geometry and boundary conditions (see, for example,~\cite{howison2005}). Doing so `factors' the dimensional units out of the problem---leaving only dimensionless equations with dimensionless parameters. The relative magnitudes of the dimensionless parameters characterize the dominant dynamics---comparable to how the Reynolds number describes fluid flow. Each process in the laser can be better understood by re-scaling the time, energies, and amplitude by the convenient scalings:
\begin{align}
	T &= T_M \widetilde{T},& E &= \Es \widetilde{E},& A &= \left( \frac{\Es}{T_M} \right)^{1/2} \widetilde{A}.
\end{align}

@@

We consider the pulse given by
\begin{align}
	A = \sqrt{P} \exp \left( -(1 + iC) \frac{T^2}{2 \sigma^2} \right) \exp(i \phi_0),
	\label{eq:A0}
\end{align}
where $P$ is the peak power, $C$ is the chirp, $\sigma^2$ is the variance, and $\phi_0$ is the phase; the pulse is shown in Fig.~\ref{fig:samplegauss}. By computing \eqref{eq:A0} after one round trip, and solving $\mathcal{L}(A) = A \textrm{e}^{i \phi}$, tedious manipulations yield the conditions
\begin{align}
	\frac{\sigma^4}{1 - \sigma^2} &= \left( \sigma^2 + 2 C s^2 \right)^2 + 4 s^4, \label{eq:sigrelation} \\
	\frac{C}{1 - \sigma^2} &= C + 2 \frac{s^2}{\sigma^2} \left( 1 + C^2 \right), \label{eq:chirprelation} \\
	1 &= \frac{W(a E_g \textrm{e}^{E_g})}{E_g} h^2 \left( 1 - \sigma^2 \right)^{1/2} \label{eq:energyrelation}
\end{align}
for \eqref{eq:A0} to be in equilibrium. Using \eqref{eq:sigrelation} and \eqref{eq:chirprelation} we eliminate the chirp, $C$, and find an expression relating $\sigma^2$ and $s$, namely
\begin{align}
	\sigma^8 + 4 s^4 \sigma^6 - 20 s^4 \sigma^4 + 32 s^4 \sigma^2 - 16 s^4 = 0.
	\label{eq:var}
\end{align}
\eqref{eq:var} has the positive, real, analytic solution
\begin{equation}
	\begin{split}
		\sigma^2 = \sqrt{2} s \left( s^6 + 3s^2 + \sqrt{4 + s^4} \left( 1 + s^4 \right) \right)^{1/2} & \\
		- s^4 - s^2 & \sqrt{4 + s^4}.
		\label{eq:equilvar}
	\end{split}
\end{equation}
We find the chirp is
\begin{align}
	C = \frac{\sigma^4 - \left( \sigma^8 - 16 s^4 \left( 1 - \sigma^2 \right)^2 \right)^{1/2}}{4 s^2 \left( 1 - \sigma^2 \right)}
	\label{eq:chirp}
\end{align}
by \eqref{eq:chirprelation}. We obtain the energy of the pulse entering the gain fibre, $E_g$, from \eqref{eq:energyrelation}:
\begin{align}
	E_g &= \frac{h^2 \left( 1 - \sigma^2 \right)^{1/2} \log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{h^2 \left( 1 - \sigma^2 \right)^{1/2} - 1},
\end{align}
and thus, the equilibrium energy at the optical coupler, is
\begin{align}
	E &= W \left( a E_g \textrm{e}^{E_g} \right), \\
	&= \frac{\log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{1 - h^2 \left( 1 - \sigma^2 \right)^{1/2}}.
	\label{eq:analenergy}
\end{align}
The argument of the logarithm in \eqref{eq:analenergy} gives us the condition
\begin{align}
	a h^2 (1 - \sigma^2)^{1/2} > 1
	\label{eq:energycond}
\end{align}
for the pulse to be sustainable.

Equating the energy of the pulse given by \eqref{eq:A0} and \eqref{eq:analenergy}, we find the peak power to be
\begin{align}
	P &= \frac{\log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{\sqrt{\pi} \sigma \left( 1 - h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}.
	\label{eq:analpower}
\end{align}

The forms of \eqref{eq:equilvar}--\eqref{eq:analpower} make the expressions difficult to work with. However, recall that $s$ is small, thus, we asymptotically expand \eqref{eq:equilvar}--\eqref{eq:analpower} assuming $s \ll 1$, and
\begin{align}
	\sigma^2 &\sim 2s(1 - s) + \bigO{s^3}, \\
	C &\sim 1 - s + \frac{1}{2}s^2 + \bigO{s^3}, \\
	E &\sim \frac{\log (a h^2)}{1 - h^2} - \frac{1}{1 - h^2} \left( 1 + \frac{h^2 \log(a h^2)}{1 - h^2}  \right) s + \bigO{s^2}, \label{eq:maxenergy} \\
	P &\sim \frac{\log(ah^2)}{\sqrt{2 \pi}(1 - h^2)} s^{-1/2} + \bigO{s^{1/2}},
\end{align}
in the limit $s \rightarrow 0$. The expansions in the limit $s \rightarrow \infty$ are given in Appendix \ref{sec:largedisp}.

@@

Fig.~\ref{fig:breakevo} highlights the phenomenon of modulation instability. By decreasing the dispersion compared to Fig.~\ref{fig:chirpft}, we find the pulse begins to `breathe'. During the first two dozen round trips of the cavity the SPM compounds and becomes detrimental to the pulse. In turn, modulation instability is induced, and degrades the pulse, until the pulse is no longer stable or sustainable. On the other hand, having a more moderate value of $b$, the pulse is able to equilibrate even with a less favourable, random, initial pulse (Fig.~\ref{fig:convevo}), as in~\cite{meng2020}. The pulse is able to shed the initial left lobe ($T \approx -1 / 2$) because of dispersion and modulation, and a central lobe ($T \approx 0$) forms, which is able to grow because of the gain medium. The central lobe then allows the pulse to come to equilibrium quickly. Furthermore, since the intensity of the initial pulse is small relative to the saturation energy, the effect of SPM is negligible during the first few round trips, allowing the laser to select preferable modes. The preferable modes then get amplified, stabilizing the pulse. Evidently, the initial shape of the pulse is much less important than the initial energy of the pulse, or the strength of the nonlinearity.
